---
title: "A297/07 Advanced R"
subtitle: "Day 3 -- Student Version (Guided Worksheet)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
editor: visual
---

This document follows the Day 3 slides. Run the code chunks together with the presentation, and complete the exercises in the marked spaces.

::: callout-note
Section 3 (strategies for writing efficient R code) is still a placeholder, waiting on the rest of the Day 3 slides. Sections 1 and 2.1--2.3 follow `Day3_TG.pptx` and `ggplot.pptx`.
:::

## Section 1. Working with missing values and factors

**Goal:**

- Detect, summarise, and handle missing values (NA) safely
- Recode and reorder categorical data using factors

### Missing values in base R

We work with the NHANES dataset (US National Health and Nutrition Examination Survey) -- demographic, examination, and lab data. `nhanes_sub` is a provided subset (see the course data files) -- load it with:

```{r}
#| label: load-packages-day3
library(dplyr)
library(forcats)
library(ggplot2)

nhanes_sub <-readRDS("~/PhD/01_Admin/Kurser_og_ECTS/06. Avanced R/minidplyr/R code/data/raw_data/nhanes_sub.RDS")
```

Before analysing a dataset: how much is missing, and where?

```{r}
nrow(nhanes_sub)
colSums(is.na(nhanes_sub))
```

Missing values are not just a technical inconvenience -- dropping every incomplete row can lose a meaningful part of the dataset, and can bias estimates if the missingness is not random. Is BMI equally likely to be missing in every age group?

```{r}
nhanes_sub %>%
  mutate(
    age_group = cut(
      Age,
      breaks = c(-Inf, 30, 45, 60, Inf),
      labels = c("<30", "31-45", "46-60", "60+")
    )
  ) %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    pct_bmi_missing = round(mean(is.na(BMI)) * 100, 1),
    .groups = "drop"
  )
```

NA propagates through calculations -- R won't guess an answer from incomplete information unless you explicitly ask it to:

```{r}
mean(nhanes_sub$BPSysAve)
mean(nhanes_sub$BPSysAve, na.rm = TRUE)
sum(nhanes_sub$BPSysAve)
sum(nhanes_sub$BPSysAve, na.rm = TRUE)
```

NA is not the only special value. `NULL` is the "undefined" value -- `is.null()` checks it, and `NULL` is silently dropped when combined into a vector:

```{r}
is.null(NULL)
my_vector <- c(25, NA, 10, NULL)
my_vector
```

`NaN`: "not a number" -- e.g. `0/0`. Check with `is.nan()`:

```{r}
0 / 0
is.nan(0 / 0)
```

`Inf`: an infinite value, e.g. dividing by 0. Check with `is.finite()`:

```{r}
1 / 0
is.finite(1 / 0)
```

`Inf` and `NaN` are **not** removed by `na.rm = TRUE` -- use `is.finite()` to subset to only finite values when that's what you need:

```{r}
z <- c(1, 22, NA, Inf, NaN, 5)
max(z, na.rm = TRUE)          # still Inf!
max(z[is.finite(z)])          # the actual largest finite value
```

"NAs introduced by coercion": an illegal conversion (e.g. text that isn't a number) silently becomes NA, with a warning:

```{r}
as.numeric(c("10", "20", "thirty", "40"))
```

NULL is ignored in a vector.

```{r}
my_vector <- c(25, NA, 10, NULL)	# define
my_vector	# print
```

```{r}
# Variance of a single number is undefined
var(22)
```

If you need to write NA explicitly for a specific type:

- `NA` -- dates or logical `TRUE`/`FALSE`
- `NA_character_` -- character columns
- `NA_real_` -- numeric columns

### Finding and removing incomplete rows

`complete.cases()`: logical vector, `TRUE` where a row has no NAs at all.

```{r}
sum(complete.cases(nhanes_sub))
sum(!complete.cases(nhanes_sub))
```

```{r}
nhanes_sub[complete.cases(nhanes_sub), ]   # keep only complete rows
nhanes_sub[!complete.cases(nhanes_sub), ]  # inspect the incomplete rows
```

`na.omit()`: a shorter way to remove incomplete rows -- but it can silently remove a lot of data, so always check how much is lost first.

```{r}
na.omit(nhanes_sub)
```

### Missing values with dplyr and naniar

Instead of checking every variable separately...

```{r}
sum(is.na(nhanes_sub$BMI))
sum(is.na(nhanes_sub$BPSysAve))
```

...apply the same operation to every column with `across()`:

```{r}
nhanes_sub %>%
  summarise(across(everything(), ~ sum(is.na(.x))))
```

Percentages instead of counts:

```{r}
nhanes_sub %>%
  summarise(across(everything(), ~ round(mean(is.na(.x)) * 100, 1)))
```

The naniar package gives the same missingness habit as one number or one plot instead of a wide table:

```{r}
# install.packages("naniar")
library(naniar)
pct_miss(nhanes_sub)
n_miss(nhanes_sub)
gg_miss_var(nhanes_sub, show_pct = TRUE)
gg_miss_var(nhanes_sub, show_pct = TRUE, facet = Gender)
```

`geom_miss_point()`: visualise missingness directly in a scatterplot -- by default `ggplot()` just drops points with a missing value.

```{r}
ggplot(data = nhanes_sub, mapping = aes(x = BMI, y = BPSysAve)) +
  geom_miss_point()
```

`gg_miss_fct()`: a heatmap of percent missingness by a factor column.

```{r}
gg_miss_fct(nhanes_sub, Education)
```

### Filling in missing values: coalesce()

`coalesce()` moves left to right and returns the first non-missing value -- e.g. the first available systolic BP reading per participant.

```{r}
nhanes_sub %>%
  mutate(bp_first_available = coalesce(BPSys1, BPSys2, BPSys3)) %>%
  select(ID, BPSys1, BPSys2, BPSys3, bp_first_available)
```

### Exercise 1.1 (5 min)

Using `nhanes_sub`, create a new variable called `BMI_demo`. For participants with an observed BMI, keep the original value. For participants with a missing BMI, replace it with the median BMI calculated from the observed values in `nhanes_sub`. Then display only: `ID`, original `BMI`, new `BMI_demo`. Use `mutate()`, `coalesce()`, and `median()`.

(Replacing missing values with the median is shown here only to demonstrate the R syntax -- it is not automatically an appropriate statistical strategy for a real analysis.)

```{r}
# Your code here
?if_else

nhanes_sub <- nhanes_sub %>% 
  mutate(BMI_demo = if_else(!is.na(BMI), BMI, mean(BMI)) 

view(nhanes_sub)

```

### Conditional transformations: case_when() and if_else()

`case_when()`: several ordered conditions, one new column.

```{r}
nhanes_sub %>%
  mutate(
    age_group = case_when(
      is.na(Age) ~ NA,
      Age < 40 ~ "under 40",
      Age < 65 ~ "40-64",
      Age >= 65 ~ "65+"
    )
  ) %>%
  select(ID, Age, age_group)
```

`case_when()` evaluates conditions top to bottom and uses the **first** one that is `TRUE` -- order matters! Put specific conditions first, and use `.default` as the final catch-all instead of spelling out the last condition explicitly:

```{r}
nhanes_sub %>%
  mutate(
    age_group = case_when(
      is.na(Age) ~ NA,
      Age < 40 ~ "under 40",
      Age < 65 ~ "40-64",
      .default = "65+"
    )
  ) %>%
  select(ID, Age, age_group)
```

`if_else()`: convenient for one binary condition.

```{r}
dplyr::if_else(nhanes_sub$BPSysAve >= 140, ">=140", "<140")
```

Missing blood-pressure values stay missing by default -- handle them explicitly with the `missing =` argument:

```{r}
dplyr::if_else(
  nhanes_sub$BPSysAve >= 140, ">=140", "<140",
  missing = "not measured"
)
```

`dplyr::if_else()` is stricter about output types than base `ifelse()` -- this catches accidental type-mixing that `ifelse()` would silently coerce instead:

```{r}
#| error: true
dplyr::if_else(nhanes_sub$BPSysAve >= 140, "high", 0)
```

### Exercise 2.1 (5 min)

Create a variable called `bmi_category` using the standard adult BMI categories:

- `"underweight"`: BMI \< 18.5
- `"normal"`: 18.5 to \<25
- `"overweight"`: 25 to \<30
- `"obese"`: BMI ≥30
- a missing BMI should remain missing

```{r}
# Your code here

```

### Factors

Factor order matters. `HealthGen` records self-reported general health, and its levels are already in a meaningful (non-alphabetical) order: Excellent -\> Vgood -\> Good -\> Fair -\> Poor.

```{r}
levels(nhanes_sub$HealthGen)
```

If you convert back to character and recreate the factor without specifying an order, R falls back to its (alphabetical) default:

```{r}
health_default <- factor(as.character(nhanes_sub$HealthGen))
levels(health_default)
```

`fct_relevel()`: set an explicit order (e.g. for a figure running from Poor -\> Excellent).

```{r}
health_ordered <- fct_relevel(
  health_default,
  "Poor", "Fair", "Good", "Vgood", "Excellent"
)
levels(health_ordered)
```

`fct_reorder()`: reorder categories by another variable, rather than a fixed order you specify by hand -- very useful in ggplot2, where categories are often ordered by the value being plotted.

```{r}
education_bmi <- nhanes_sub %>%
  filter(!is.na(Education), !is.na(BMI)) %>%
  group_by(Education) %>%
  summarise(mean_bmi = mean(BMI))
education_bmi

education_bmi <- education_bmi %>%
  mutate(Education = fct_reorder(Education, mean_bmi))
levels(education_bmi$Education)
```

`fct_relevel()` = you already know the order you want. `fct_reorder()` = the order should come from another variable.

`fct_recode()`: rename a level, without changing the underlying category.

```{r}
fct_recode(nhanes_sub$MaritalStatus, "Partnered" = "LivePartner")
```

`fct_collapse()`: combine several levels into one -- only do this when the grouping is substantively appropriate, since it changes what the variable represents.

```{r}
fct_collapse(
  nhanes_sub$MaritalStatus,
  NotMarried = c("Divorced", "Separated", "Widowed", "NeverMarried")
)
```

`fct_lump_n()`: keep the n most frequent levels, combine the rest into "Other".

```{r}
marital_full <- nhanes_sub %>%
  filter(!is.na(MaritalStatus)) %>%
  pull(MaritalStatus)
fct_count(marital_full, sort = TRUE)
fct_count(fct_lump_n(marital_full, n = 3), sort = TRUE)
```

`fct_infreq()`: order levels from most common to least common.

```{r}
fct_count(fct_infreq(marital_full))
```

## Section 2.1. Data visualization with ggplot2 I: choosing plots and core grammar

**Goal:**

- Understand the grammar of graphics: data, `aes()`, and geoms
- Build a basic plot and map variables to aesthetics

### Walkthrough (follow along with the slides)

ggplot is a package for making plots -- compared to base R plotting, you can customize plots much more easily.

All ggplots can be built from variations of this basic form:

```         
ggplot(data = <data>) + <geom_function>(mapping = aes(<mappings>))
```

There are countless options -- the cheat sheet is your friend: <https://rstudio.github.io/cheatsheets/data-visualization.pdf>

```{r}
library(ggplot2)

# We use the iris dataset already built into R (most online ggplot
# tutorials use it too, so it's worth getting used to)
iris
```

```{r}
# geom_point(): a basic scatterplot
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width))
```

```{r}
# Add colors based on a categorical variable: col/color
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species))
```

```{r}
# You can add more than one geom_function
ggplot(data = iris) +
  geom_point(mapping = aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_line(mapping = aes(x = Petal.Length, y = Petal.Width, color = Species))
```

```{r}
# Avoid redundancy: data and aes() shared in ggplot() apply to every geom
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point() +
  geom_line()
```

```{r}
# Change appearance of items: shape, size, lwd (line width)
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species, shape = Species)) +
  geom_point(size = 3) +
  geom_line(lwd = 1)
```

```{r}
# Grouping: without a group, geom_line() connects ALL points in one line
ggplot(iris, aes(Petal.Length, Petal.Width)) +
  geom_point(aes(col = Species)) +
  geom_line()
```

```{r}
# Grouping: group draws a separate line per Species
ggplot(iris, aes(Petal.Length, Petal.Width)) +
  geom_point(aes(col = Species)) +
  geom_line(aes(group = Species))
```

```{r}
# Layers: be aware of the order! Later layers are drawn on top of earlier ones
ggplot(iris, aes(Petal.Length, Petal.Width)) +
  geom_line(aes(group = Species)) +
  geom_point(aes(col = Species))
```

### Exercise (10 min)

Try to recreate the plot shown on the slide: a scatterplot of `Petal.Length` (x) vs `Petal.Width` (y) from `iris`, with points colored **and** shaped by `Species`, drawn a bit larger and semi-transparent.

```{r}
# Your code here

```

## Section 2.2. Data visualization with ggplot2 II: faceting, themes, and labels

**Goal:**

- Split plots into panels with `facet_wrap()`/`facet_grid()`
- Adjust themes and axis/plot labels

### Walkthrough (follow along with the slides)

`facet_grid()`: split a plot into panels by a variable.

```{r}
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point() +
  facet_grid(cols = vars(Species))
```

`facet_grid()` vs `facet_wrap()`: `facet_grid()` will still display a panel for combinations that are empty; `facet_wrap()` will not.

```{r}
iris_extra <- iris %>%
  mutate(is_pretty = sample(c("ugly", "pretty"), nrow(iris), replace = TRUE))

ggplot(iris_extra, aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point() +
  facet_grid(Species ~ is_pretty)

ggplot(iris_extra, aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point() +
  facet_wrap(Species ~ is_pretty, nrow = 3)
```

::: callout-note
**Practical point:** you can often avoid having to combine plots afterwards if you facet cleverly instead. patchwork (used later today) is great for combining a handful of different plots with annotations (A, B, ...), but for grids of many similar plots (e.g. simulation studies), facets are usually the better choice.
:::

```{r}
# Scaling of axes
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species, shape = Species)) +
  geom_point(size = 3) +
  scale_x_log10(breaks = 1:7)
```

```{r}
# Themes and titles
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point() +
  theme_bw() +
  labs(
    title = "Petal width against petal length",
    x = "Petal length",
    y = "Petal width"
  )
```

There are many built-in themes beyond `theme_bw()` -- see <https://statisticsglobe.com/ggplot2-themes-r>.

## Section 2.3. Polishing figures: scales, annotations, and publication-ready plots

**Goal:**

- Adjust scales, add annotations, and combine plots
- Save figures in a publication-ready format

### Walkthrough (follow along with the slides)

Example dataset for a forest plot (simulated):

```{r}
df <- data.frame(
  estimate = runif(6, 0.8, 2.5),
  disorder = factor(rep(c("Schizophrenia", "Depression", "ADHD"), 2)),
  sex = factor(c(rep("Male", 3), rep("Female", 3)))
) %>%
  rowwise() %>%
  mutate(
    upper = runif(1, estimate, estimate + runif(1, 0, 1)),
    lower = runif(1, estimate - runif(1, 0, 0.8), estimate)
  )
```

Forest plot: `geom_point()` + `geom_errorbar()`, dodged so male/female don't overlap.

```{r}
ggplot(df, aes(estimate, disorder, col = sex)) +
  geom_point(position = position_dodge(width = 0.2)) +
  geom_errorbar(
    aes(xmin = lower, xmax = upper),
    width = 0.1,
    position = position_dodge(width = 0.2)
  ) +
  theme_bw() +
  labs(x = "HR (95% CI)", y = NULL)
```

The same plot, using a facet instead of dodging:

```{r}
ggplot(df, aes(estimate, disorder, col = sex)) +
  geom_point() +
  geom_errorbar(aes(xmin = lower, xmax = upper), width = 0.1) +
  facet_grid(cols = vars(sex)) +
  theme_bw() +
  labs(x = "HR (95% CI)", y = NULL)
```

Combining plots: the patchwork package.

```{r}
# install.packages("patchwork")
library(patchwork)
```

- `plot1 | plot2` -- plots side by side (can also use `+`)
- `plot1 / plot2` -- plots stacked on top of each other
- `plot_annotation()` -- annotate the combined figure (e.g. a title)
- `plot_layout()` -- control layout, guides, and axes
- `theme()` also works on a combined patchwork, just like in ggplot

See <https://patchwork.data-imaginist.com/index.html>.

`geom_text()`/`geom_label()`: labels.

```{r}
ggplot(iris, aes(Petal.Width, Petal.Length,
                 label = paste(Petal.Width, Petal.Length, sep = ", "))) +
  geom_point() +
  geom_text()

ggplot(iris, aes(Petal.Width, Petal.Length,
                 label = paste(Petal.Width, Petal.Length, sep = ", "))) +
  geom_point() +
  geom_label()
```

`geom_abline()`/`geom_hline()`/`geom_vline()`: reference lines -- e.g. a dashed line at HR = 1 on the forest plot above.

```{r}
ggplot(df, aes(estimate, disorder, col = sex)) +
  geom_point(position = position_dodge(width = 0.2)) +
  geom_errorbar(
    aes(xmin = lower, xmax = upper),
    width = 0.1,
    position = position_dodge(width = 0.2)
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  theme_bw() +
  labs(x = "HR (95% CI)", y = NULL)
```

`theme()`: adjusting the theme in general (beyond a built-in `theme_*()`).

```{r}
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species, shape = Species)) +
  geom_point(size = 3, alpha = 0.6) +
  labs(title = "Scatterplot") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom",
    axis.title = element_text(size = 15)
  )
```

`ggsave()`: saving a figure.

```{r}
#| eval: false
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species, shape = Species)) +
  geom_point(size = 3, alpha = 0.6)

ggsave("save_figure.pdf", dpi = 600, width = 10, height = 10)
```

It's impossible to remember every customization option -- googling is completely normal. This session is not an exhaustive list of what ggplot2 can do.

A few more useful geoms for later:

- `geom_point()` -- points
- `geom_errorbar()` -- bars (useful for CIs with `geom_point()`)
- `geom_text()` -- text labels on a plot
- `geom_line()` -- lines
- `geom_ribbon()` -- ribbons (useful for CIs with `geom_line()`)
- `geom_histogram()` -- histogram
- `geom_density()` -- density
- `geom_smooth()` -- smoothed trend line
- `geom_boxplot()` -- boxplot
- `geom_abline`/`hline`/`vline()` -- reference lines (e.g. HR = 1)

## Section 3. Strategies for writing efficient R code

**Goal:**

- Recognise common performance pitfalls
- Apply simple strategies to write faster, cleaner R code

### Walkthrough (follow along with the slides)

```{r}


```

### Exercise

```{r}


```

## Section 4. Guided practice: wrangling, visualization, and code efficiency

**Goal:** Apply Day 3's ggplot2 skills in a single guided mini project.

### Practice

Today's guided practice is a standalone mini project focused on ggplot2 -- missing values/factors and joining/wrangling already got a combined mini project on Day 2, and code efficiency (Section 3) doesn't have a natural single dataset to combine everything into, so this one goes deep on visualization instead, building on the `mpg` exercise from the ggplot session.

Open the separate project files instead of working in this document:

- `Day 3 mini project - exercise.R` (work through this one)
- `Day 3 mini project - exercise solutions.R` (only check this **after** attempting the exercise yourself)

## End-of-day reflection

1.  One thing I understood well today:

    ------------------------------------------------------------------------

2.  One thing that is still unclear:

    ------------------------------------------------------------------------

3.  One command or concept I think I will use again:

    ------------------------------------------------------------------------
